---
title: "Bora"
output: html_document
---
This project explores the properties and frequency of the Bora wind in Trieste based on hourly meteorological observations of the Trieste molo F.lli Bandiera observatory, provided by OSMER. The Bora is herein assumed to be any wind coming from the north (N), northeast (NE), or east (E) with a speed of at least 10 m/s (36 km/h).

We will explore several aspects of the wind: 
- How often does the Bora blow in Trieste?
- Monthly patterns to observe seasonal behavior.
- Is it true, as people claim in Trieste, that the Bora blows for an odd number of days?
- Annual trends of the Bora (by month)
- Daily trends of the Bora (by hour).
- Historical trends of the Bora (over time).
- A specific measurement in the data set (temperature, pressure, etc.) and its association with the Bora.

This work provides insight into the dynamics of one of Trieste's most iconic meteorological phenomena.

We took wind data taken in Trieste molo F.lli Bandiera. We put it all together, using the semicolon as a delimiter and taking the column headers from the first row. Additionally, we added a YEAR column for the appended dataset that contains all years. We also changed the column names from Italian to English. Missing values marked as "-" in the dataset were changed to NA because NA is a standard in R for representing missing or unavailable data. Leaving the value as NA ensures that it won’t be accidentally treated as a valid number in calculations or visualizations. We filtered the Bora wind out of all winds recorded in Trieste by these thresholds: speed of at least 36 km/h, coming from N (North) 0°, NE (Northeast) 45°, ENE (East-Northeast) 67.5°, or E (East) 90°. 



```{r}
library(tidyverse)

# Taking data from 2000-2024, from Trieste molo F.lli Bandiera.

# The CSVs use semicolon (;) as the delimiter. They have column headers on the first row (only once per file). They need to be combined without duplicating headers. A new YEAR column needs to be added from the filename.

# all the CSV files - read with a regex that selects filenames named "exactly 4 digits.csv"

files <- list.files("data", pattern = "^\\d{4}\\.csv$", full.names = TRUE)
 
# Function to read a single CSV and add YEAR column to distinguish between years
read_with_year <- function(file) {
  year <- str_extract(basename(file), "\\d{4}") # add year from the filename

  df <- suppressWarnings(
    read_delim(file, delim = ";", col_types = cols(.default = "c"), na = "-") 
  )
  
  df <- df |> mutate(YEAR = year)
  return(df)
}

# Read and combine all
all_data <- map_dfr(files, read_with_year)

# Move YEAR column to the front, next to month, day, hour.
all_data <- all_data |> relocate(YEAR)

# Convert all columns EXCEPT YEAR to numeric if possible  -  NA WHERE DATA IS MISSING
# Some of the years have all "-" in some columns => NA.
# Reason - NA is a standard in R for representing missing or unavailable data. By leaving the value as NA, it ensures that it won’t be accidentally treated as a valid number in calculations or visualizations.

all_data <- all_data |>
  mutate(across(.cols = -YEAR, ~ suppressWarnings(as.numeric(.))))

# Check result
head(all_data)


# This renames the column names to English for easier handling.

all_data <- all_data |>
   rename(
   wind_speed_kmH = `Vento med km/h`, 
   wind_direction_FromN = `Direzione Vento gradi N`,
    pressure = `Pressione hPa`,
    month = `mese`,
    day = `giorno`,
    hour = `ora UTC*`,
    day = `giorno`,
   rainMM = `Pioggia mm`,
   temperatureC = `Temp. gradi C`,
   humidity = `Umidita' %`,
  wind_speed_kmH_MAX = `Vento max km/h`,
   radiation_KJ_m2 = `Radiaz. KJ/m2`,
   pressure_hPa = `Pressione hPa`,
   leaf_wetness_min = `Bagnatura Fogliare min.`
 )

head(all_data)

# Filter Bora events
bora_data <- all_data |>
  filter(
    wind_direction_FromN >= 0 & wind_direction_FromN <= 90,
    wind_speed_kmH >= 36
  )


head(bora_data)
```
## How often does the Bora blow in Trieste?

Firstly, we will analyze the occurences of Bora in Trieste per year.

```{r}
# Convert again after renaming
all_data <- all_data |>
  mutate(across(c(wind_speed_kmH, wind_direction_FromN, pressure_hPa), as.numeric))


# Yearly count of Bora events
bora_by_year <- bora_data |>
  group_by(YEAR) |>
  summarise(events = n(), .groups = "drop")

# Plot
ggplot(bora_by_year, aes(x = as.numeric(YEAR), y = events)) +
  geom_col(fill = "lightpink") +
  labs(
    title = "Bora Events per Year in Trieste (2000–2024)",
    x = "Year",
    y = "Number of Events"
  ) +
  theme_minimal()

```
Then, we're going to analyze how often the Bora wind appears per month, by taking the average across the years in the dataset. The Bora winds seem to occur more frequently during the colder months, from autumn to early spring (October through March).
```{r}
library(tidyverse)

# First, ensure 'mese' is numeric or integer (in case it's in a different format)
bora_data$month <- as.integer(bora_data$month)

# Check if there are any missing values in 'month'
sum(is.na(bora_data$month))  # This will give you the number of missing values

# Remove rows with NA in month (if any)
bora_data <- bora_data |>
  filter(!is.na(month))

# Count Bora events per month
bora_by_month <- bora_data |>
  group_by(month) |>
  summarise(events = n(), .groups = 'drop')

# Convert 'month' to a proper factor for correct ordering in the plot
bora_by_month$month <- factor(bora_by_month$month, levels = 1:12, labels = month.name)

# Plot the Bora events by month
ggplot(bora_by_month, aes(x = month, y = events)) +
  geom_col(fill = "purple") +
  labs(
    title = "Number of Bora Events per Month in Trieste (2000–2024)",
    x = "Month",
    y = "Number of Events"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## 2. Is it true, as people claim in Trieste, that the Bora blows for an odd number of days?
Bora blew for 997 streaks in the record  
605 had an odd length while 392 had an even length  
The exact binomial test gives a p value of 1.6 × 10⁻¹¹ which is far below 0.05 so the result is significant  
The estimated chance that any streak has an odd length is about 0.61 with a 95 percent confidence range from 0.58 to 0.64  
The data back up the local belief that the Bora tends to blow for an odd number of days
```{r}
library(tidyverse)

# unique Bora days
bora_days <- bora_data |>
  mutate(date = as.Date(paste(YEAR, month, day, sep = "-"))) |>
  distinct(date) |>
  arrange(date)

# consecutive‑day streak ids
bora_days <- bora_days |>
  mutate(new_streak = c(TRUE, diff(date) != 1)) |>
  mutate(streak_id = cumsum(new_streak))

# length of each streak
streaks <- bora_days |>
  group_by(streak_id) |>
  summarise(length = n(), .groups = "drop")

# counts of odd vs even streaks
odd_count  <- sum(streaks$length %% 2 == 1)
even_count <- sum(streaks$length %% 2 == 0)
total_streaks <- nrow(streaks)

# exact binomial test
binom.test(odd_count, total_streaks, p = 0.5)

# bar plot
ggplot(tibble(parity = c("Odd", "Even"),
              count  = c(odd_count, even_count)),
       aes(parity, count, fill = parity)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = count), vjust = -0.3) +
  scale_fill_manual(values = c("Odd" = "lightpink", "Even" = "grey")) +
  labs(title = "Bora streaks by parity of length",
       x = NULL, y = "Number of streaks") +
  theme_minimal()

```

## 3. Average monthly distribution (seasonality)
We used hourly wind data from Trieste molo Flli Bandiera for the years 2000 to 2024. A Bora hour is any record with wind speed at least thirty six kilometres per hour and a direction between zero and ninety degrees measured clockwise from north. First all hourly files were merged and the numeric fields were cleaned. Then a filter kept only the Bora hours. For seasonality we grouped these Bora hours by year and by numeric month. That gave twelve counts for every year. Next we took the mean of those counts across the twenty five year span to obtain the average number of Bora events for each month. The tibble printed below the graph lists those averages. January shows about eighty seven events per year. February is even higher near ninety one. Counts then fall through spring and reach a minimum near twenty seven in June. Activity begins to rise in August and climbs back above seventy in November with December near seventy eight. The bar plot visualises this curve. Each light blue bar stands for a month numbered one through twelve on the horizontal axis. The vertical axis shows the mean count of hourly events. The tallest bars sit over the winter months while the lowest bar sits over June. The shape confirms that the Bora is most active from mid autumn through late winter and quietest in early summer.
```{r bora_monthly_avg}
bora_by_month_year <- bora_data |>
  group_by(YEAR, month) |>
  summarise(events = n(), .groups = "drop")

bora_month_avg <- bora_by_month_year |>
  group_by(month) |>
  summarise(avg_events = mean(events), .groups = "drop") |>
  mutate(month = factor(month, levels = 1:12))

ggplot(bora_month_avg, aes(month, avg_events)) +
  geom_col(fill = "lightblue") +
  labs(title = "Average monthly Bora events (2000–2024)", x = "Month", y = "Avg number of events") +
  theme_minimal()

# print table for report
print(bora_month_avg)
```


## 4. Historical trend (total events per year)
We kept only the hours when wind speed was at least thirty six kilometres per hour and the direction was between north and east. Then we counted those Bora hours for every year from two thousand to two thousand twenty four. The counts swing widely each year from about four hundred to almost nine hundred events. A linear model on these totals gives a flat slope and a high p value so time explains almost none of the variation. The result says that Bora activity shows no clear upward or downward trend over the last twenty five years even though single years can be very calm or very busy.
```{r bora_year_trend}
bora_year_total <- bora_data |>
  group_by(YEAR) |>
  summarise(total_events = n(), .groups = "drop") |>
  mutate(YEAR = as.numeric(YEAR))

ggplot(bora_year_total, aes(YEAR, total_events)) +
  geom_line(linewidth = 0.8, colour = "steelblue") +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed", linewidth = 0.6) +
  labs(title = "Annual Bora events in Trieste (2000–2024)", x = "Year", y = "Number of events") +
  theme_minimal()

lm_res <- lm(total_events ~ YEAR, data = bora_year_total)
summary(lm_res)

# print yearly totals for reference
print(bora_year_total)
```