---
title: "Bora"
output: html_document
---
# Bora

This project explores the properties and frequency of the Bora wind in Trieste based on hourly meteorological observations of the Trieste molo F.lli Bandiera observatory, provided by OSMER. The Bora is herein assumed to be any wind coming from the north (N), northeast (NE), or east (E) with a speed of at least 10 m/s (36 km/h).

We will explore several aspects of the wind: 

- How often does the Bora blow in Trieste?
- Monthly patterns to observe seasonal behavior.
- Is it true, as people claim in Trieste, that the Bora blows for an odd number of days?
- Annual trends of the Bora (by month)
- Daily trends of the Bora (by hour).
- Historical trends of the Bora (over time).
- A specific measurement in the data set (temperature, pressure, etc.) and its association with the Bora.

This work provides insight into the dynamics of one of Trieste's most iconic meteorological phenomena.

We took wind data taken in Trieste molo F.lli Bandiera. We put it all together, using the semicolon as a delimiter and taking the column headers from the first row. Additionally, we added a YEAR column for the appended dataset that contains all years. We also changed the column names from Italian to English. Missing values marked as "-" in the dataset were changed to NA because NA is a standard in R for representing missing or unavailable data. Leaving the value as NA ensures that it won’t be accidentally treated as a valid number in calculations or visualizations. We filtered the Bora wind out of all winds recorded in Trieste by these thresholds: speed of at least 36 km/h, coming from N (North) 0°, NE (Northeast) 45°, ENE (East-Northeast) 67.5°, or E (East) 90°. 



```{r}
library(tidyverse)

# Taking data from 2000-2024, from Trieste molo F.lli Bandiera.

# The CSVs use semicolon (;) as the delimiter. They have column headers on the first row (only once per file). They need to be combined without duplicating headers. A new YEAR column needs to be added from the filename.

# all the CSV files - read with a regex that selects filenames named "exactly 4 digits.csv"

files <- list.files("data", pattern = "^\\d{4}\\.csv$", full.names = TRUE)
 
# Function to read a single CSV and add YEAR column to distinguish between years
read_with_year <- function(file) {
  year <- str_extract(basename(file), "\\d{4}") # add year from the filename

  df <- suppressWarnings(
    read_delim(file, delim = ";", col_types = cols(.default = "c"), na = "-") 
  )
  
  df <- df |> mutate(YEAR = year)
  return(df)
}

# Read and combine all
all_data <- map_dfr(files, read_with_year)

# Move YEAR column to the front, next to month, day, hour.
all_data <- all_data |> relocate(YEAR)

# Convert all columns EXCEPT YEAR to numeric if possible  -  NA WHERE DATA IS MISSING
# Some of the years have all "-" in some columns => NA.
# Reason - NA is a standard in R for representing missing or unavailable data. By leaving the value as NA, it ensures that it won’t be accidentally treated as a valid number in calculations or visualizations.

all_data <- all_data |>
  mutate(across(.cols = -YEAR, ~ suppressWarnings(as.numeric(.))))

# Check result
head(all_data)


# This renames the column names to English for easier handling.

all_data <- all_data |>
   rename(
   wind_speed_kmH = `Vento med km/h`, 
   wind_direction_FromN = `Direzione Vento gradi N`,
    pressure = `Pressione hPa`,
    month = `mese`,
    day = `giorno`,
    hour = `ora UTC*`,
    day = `giorno`,
   rainMM = `Pioggia mm`,
   temperatureC = `Temp. gradi C`,
   humidity = `Umidita' %`,
  wind_speed_kmH_MAX = `Vento max km/h`,
   radiation_KJ_m2 = `Radiaz. KJ/m2`,
   pressure_hPa = `Pressione hPa`,
   leaf_wetness_min = `Bagnatura Fogliare min.`
 )

head(all_data)

# Filter Bora events
bora_data <- all_data |>
  filter(
    wind_direction_FromN >= 0 & wind_direction_FromN <= 90,
    wind_speed_kmH >= 36
  )


print(bora_data)
```
## 1. How often does the Bora blow in Trieste?

Firstly, we will analyze the occurences of Bora in Trieste per year.

```{r}
# Convert again after renaming
all_data <- all_data |>
  mutate(across(c(wind_speed_kmH, wind_direction_FromN, pressure_hPa), as.numeric))


# Yearly count of Bora events
bora_by_year <- bora_data |>
  group_by(YEAR) |>
  summarise(events = n(), .groups = "drop")

ggplot(bora_by_year, aes(x = as.numeric(YEAR), y = events)) +
  geom_col(fill = "lightpink") +
  labs(
    title = "Bora Events per Year in Trieste (2000–2024)",
    x = "Year",
    y = "Number of Events"
  ) +
  theme_minimal()

```
Then, we're going to analyze how often the Bora wind appears per month, by taking the average across the years in the dataset. The Bora winds seem to occur more frequently during the colder months, from autumn to early spring (October through March).

```{r}
library(tidyverse)

# First, ensuring 'month' is numeric or integer (in case it's in a different format)
bora_data$month <- as.integer(bora_data$month)

# Checking if there are any missing values in 'month'
sum(is.na(bora_data$month))  

# Removing rows with NA in month (if any)
bora_data <- bora_data |>
  filter(!is.na(month))

# Counting Bora events per month
bora_by_month <- bora_data |>
  group_by(month) |>
  summarise(events = n(), .groups = 'drop')

# Converting 'month' to a proper factor for correct ordering in the plot
bora_by_month$month <- factor(bora_by_month$month, levels = 1:12, labels = month.name)

# Plotting the Bora events by month
ggplot(bora_by_month, aes(x = month, y = events)) +
  geom_col(fill = "purple") +
  labs(
    title = "Number of Bora Events per Month in Trieste (2000–2024)",
    x = "Month",
    y = "Number of Events"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## 2. Is it true, as people claim in Trieste, that the Bora blows for an odd number of days?

```{r}
library(tidyverse)
library(lubridate)

bora_days <- all_data %>% 
  filter(wind_speed_kmH >= 36,
         wind_direction_FromN >= 0,
         wind_direction_FromN <= 90) %>% 
  mutate(date = as.Date(paste(YEAR, month, day, sep = "-"))) %>% 
  distinct(date) %>% 
  arrange(date) %>% 
  mutate(new_streak = c(TRUE, diff(date) > 1),
         streak_id   = cumsum(new_streak))

# length for every streak and the year it starts
streak_info <- bora_days %>% 
  group_by(streak_id) %>% 
  summarise(start_date = first(date),
            len        = n(),
            .groups    = "drop") %>% 
  mutate(year = year(start_date))

per_year_len <- streak_info %>% 
  count(year, len, name = "n_streaks")

avg_len <- per_year_len %>% 
  group_by(len) %>% 
  summarise(avg_per_year = mean(n_streaks),
            .groups = "drop")
print(avg_len)

avg_len %>% 
  filter(len <= 7) %>% 
  ggplot(aes(factor(len), avg_per_year)) +
  geom_col(fill = "steelblue") +
  labs(title = "Average Bora streak count per year by length",
       x = "Streak length in days",
       y = "Average count per year") +
  theme_minimal()

streak_info %>% 
  filter(len > 1) %>% 
  mutate(parity = if_else(len %% 2 == 0, "even", "odd")) %>% 
  count(parity)

```
## 3. Average monthly distribution (seasonality)
For seasonality we grouped these Bora hours by year and by numeric month. That gave twelve counts for every year. Next we took the mean of those counts across the twenty five year span to obtain the average number of Bora events for each month. The tibble printed below the graph lists those averages. January shows about 87 events per year. February is even higher than that, with almost 91 events. Numbers drop in the spring months and reach a minimum, that is 21 events, in June. Activity begins to rise in August and climbs back above 70 in November with December near 78. The bar plot visualises this curve. Each light blue bar stands for a month numbered one through twelve on the horizontal axis. The vertical axis shows the mean count of hourly events. The tallest bars sit over the winter months while the lowest bar sits over June. The shape confirms that the Bora is most active from mid autumn through late winter and quiets down in early summer.
```{r bora_monthly_avg}
bora_by_month_year <- bora_data |>
  group_by(YEAR, month) |>
  summarise(events = n(), .groups = "drop")

head(bora_by_month_year)

bora_month_avg <- bora_by_month_year |>
  group_by(month) |>
  summarise(avg_events = mean(events), .groups = "drop") |>
  mutate(month = factor(month, levels = 1:12))

print(bora_month_avg)

ggplot(bora_month_avg, aes(month, avg_events)) +
  geom_col(fill = "lightblue") +
  labs(title = "Average monthly Bora events (2000–2024)", x = "Month", y = "Avg number of events") +
  theme_minimal()

# print table for report
print(bora_month_avg)
```


## 4. Explore the daily trends of the Bora (by hour).
We converted UTC timestamps to Europe/Rome time. In the code we count how many events occur in each hour and fill missing hours with zero.
As we can observe on the plot daily trends of Bora peak at around 9h, and then drop to a low at 15h. They climb again in the late afternoon and evening. 


```{r bora_year_trend}
library(lubridate)
library(dplyr)

bora_local <- bora_data %>%
  mutate(
    dt_utc    = make_datetime(as.integer(YEAR), month, day, hour, tz = "UTC"),
    dt_local  = with_tz(dt_utc, "Europe/Rome"),
    local_hour = hour(dt_local)
  )

bora_by_local_hour <- bora_local %>%
  group_by(local_hour) %>%
  summarise(events = n(), .groups = "drop") %>%
  complete(local_hour = 0:23, fill = list(events = 0)) %>%
  mutate(local_hour = factor(local_hour, levels = 0:23))

print(bora_by_local_hour)

ggplot(bora_by_local_hour, aes(local_hour, events)) +
  geom_col(fill = "lightpink") +
  labs(
    title = "Total Bora Events by Local Hour (CET/CEST)",
    x     = "Local Hour",
    y     = "Number of Events"
  ) +
  theme_minimal()

```


## 5. Exploring the Historical Trends of the Bora (Over Time)

For each year, we counted how many such Bora days occurred. The result is a time series plot displaying the number of Bora days annually from 2000 to 2024.
From the graph, we observe that the number of Bora days per year fluctuates significantly, ranging from around 40 to more than 140. Despite these yearly variations, the overall trend appears relatively stable with no strong upward or downward movement. This suggests that, although some years experience more Bora activity than others, there is no significant long-term increase or decrease in the frequency of Bora days over this 25-year span.
```{r}

bora_days <- all_data |>
  filter(wind_speed_kmH_MAX * 0.27778 >= 10,
         wind_direction_FromN >= 45,
         wind_direction_FromN <= 90) |>
  mutate(date = as.Date(paste(YEAR, month, day, sep = "-"))) |>
  distinct(YEAR, date) |>
  group_by(YEAR) |>
  summarise(bora_day_count = n(), .groups = "drop")

# Plot
ggplot(bora_days, aes(x = as.numeric(YEAR), y = bora_day_count)) +
  geom_line(color = "darkblue") +
  geom_point(size = 2) +
  labs(
    title = "Number of Bora Days per Year (2000–2024)",
    x = "Year", y = "Number of Bora Days"
  ) +
  theme_minimal()
```


### 6. Finally, choose at least a different measurement available in the data set (temperature, pressure, ...) and evaluate the association with the Bora.

We calculated the average monthly temperature for both Bora and Non-Bora data. The results showed that the mean temperature on Bora days was noticeably lower than on non-Bora days. This aligns with local expectations, as Bora winds are known to bring cold, dry air from the northeast.
To visualize the difference, we used a histogram for comparing the temperature distributions, per month. The box for Bora days lies visibly lower than that for non-Bora days, reinforcing the finding that Bora is typically associated with colder weather. This result highlights a clear meteorological impact of the Bora beyond just wind behavior.

```{r}
library(tidyverse)

# Filter and define Bora and Non-Bora
temp_df <- all_data |>
  filter(!is.na(wind_speed_kmH_MAX),
         !is.na(wind_direction_FromN),
         !is.na(temperatureC),
         !is.na(month)) |>
  mutate(is_bora = wind_speed_kmH_MAX * 0.27778 >= 10 &
           wind_direction_FromN >= 45 &
           wind_direction_FromN <= 90)

# Prepare temperature dataset
temp_month_df <- temp_df |>
  mutate(
    month = factor(month, levels = 1:12, labels = month.name),
    Type = ifelse(is_bora, "Bora", "Non-Bora")
  ) |>
  group_by(month, Type) |>
  summarise(avg_temp = mean(temperatureC, na.rm = TRUE), .groups = "drop")

# Plot
ggplot(temp_month_df, aes(x = month, y = avg_temp, fill = Type)) +
  geom_col(position = position_dodge(width = 0.6), width = 0.5) +
  labs(
    title = "Average Temperature per Month: Bora vs Non-Bora Days",
    x = "Month",
    y = "Average Temperature (°C)",
    fill = NULL
  ) +
  scale_fill_manual(values = c("Bora" = "skyblue", "Non-Bora" = "orange")) +
  scale_y_continuous(breaks = seq(-10, 30, by = 5)) +  # <--- Add finer Y axis labels
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  )

```

